Towards Practical First-Order Model Counting

### 2025-01-18 doc/paper/paper.tex
Start rewriting preliminaries

\todo[inline,caption={}]{
  \begin{itemize}
    \item Don't assume that the reader knows FOMC and FOMC algorithms
    \item It would be nice to give the necessary introduction to FOKC, or
          \textsc{Crane} (an intuitive explanation early in the paper)
    \item The principal motivation of a knowledge compilation-based approach to
          FOMC comes from its ability to find polynomial-time solutions, where
          the polynomial has a lower degree compared to, e.g.,
          \textsc{FastWFOMC}~\cite{DBLP:conf/kr/DilkasB23}. The ability of the
          algorithm to mathematically describe recursive solutions is also
          noteworthy as the only other available alternative is to guess
          them~\cite{DBLP:conf/ilp/BarvinekB0ZK21}.
  \end{itemize}
}

FOMC algorithms are diverse, with approaches ranging from \emph{first-order
  knowledge compilation} (FOKC) to local
search~\cite{DBLP:journals/pvldb/NiuRDS11}, Monte Carlo
sampling~\cite{DBLP:journals/cacm/GogateD16}, and anytime
approximation~\cite{DBLP:conf/ijcai/BremenK20}. Among these, FOKC-based
algorithms are particularly prominent, transforming formulas into structured
representations such as circuits or graphs. Notable examples include
\textsc{ForcLift}~\cite{DBLP:conf/ijcai/BroeckTMDR11} and its successor
\textsc{Crane}~\cite{DBLP:conf/kr/DilkasB23}. Another important algorithm,
\textsc{FastWFOMC}~\cite{DBLP:conf/uai/BremenK21}, uses cell enumeration as its
foundation.

@@

@@

\begin{example}
  Skolemization transforms
  \begin{equation}\label[formula]{eq:skolemizationinitial}
    \forall x \in \Gamma\text{. }\exists y \in \Delta\text{. } P(x, y)
  \end{equation}
  into
  \begin{equation}\label[formula]{eq:skolemization}
    \begin{aligned}
      (\forall x \in \Gamma\text{. } &Z(x))\land{}\\
      (\forall x \in \Gamma\text{. } \forall y \in \Delta\text{. } &Z(x) \lor \neg P(x, y))\land{}\\
      (\forall x \in \Gamma\text{. } &S(x) \lor Z(x))\land{}\\
      (\forall x \in \Gamma\text{. } \forall y \in \Delta\text{. } &S(x) \lor \neg P(x, y)).
    \end{aligned}
  \end{equation}
  We will see how, with suitable weights on the new predicates $S/1$ and $Z/1$,
  the WFOMC of \cref{eq:skolemization} is equal to the FOMC of
  \cref{eq:skolemizationinitial}.
\end{example}

\todo[inline]{Make sure to revisit this in the future}

\subsection{FOMC Algorithms and Their Logics}\label{sec:threelogics}

\todo[inline]{Shorten the part about the three logics}

\begin{table*}[t]
  \centering
  \begin{tabular}{llclll}
    \toprule
    Logic & Sorts & Constants & Variables & Quantifiers & Additional atoms\\
    \midrule
    \FO & one or more & \cmark & unlimited & $\forall$, $\exists$ & $x = y$\\
    \Ctwo & one & \xmark & two & $\forall$, $\exists$, $\exists^{= k}$, $\exists^{\le k}$, $\exists^{\ge k}$ & ---\\
    \UFO & one & \xmark & two & $\forall$ & $|P| = m$\\
    \bottomrule
  \end{tabular}
  \caption{A comparison of the three logics used in FOMC\@. The
    2\textsuperscript{nd}--5\textsuperscript{th} columns refer to: the number of
    sorts, support for constants, the maximum number of variables, and supported
    quantifiers, respectively. The last column lists supported atoms in addition
    to those of the form $P(\mathbf{t})$ for a predicate $P/n$ and an $n$-tuple
    of terms $\mathbf{t}$. Here: $k$ and $m$ are non-negative integers, with the
    latter depending on the domain size, $P$ represents a predicate, and $x$ and
    $y$ are terms. }\label{tbl:logics}
\end{table*}

@@

In \cref{tbl:logics}, we outline the differences among three first-order logics
commonly used in FOMC: \FO{}, \Ctwo{}, and \UFO{}. First, \FO{} is the input
format for
\textsc{ForcLift}\footnote{\url{https://github.com/UCLA-StarAI/Forclift}} and
its extensions
\textsc{Crane}\footnote{\url{https://doi.org/10.5281/zenodo.8004077}} and
\Cranetwo{}. Second, \Ctwo{} is often used in the literature on
\textsc{FastWFOMC} and related
methods~\cite{DBLP:journals/jair/Kuzelka21,DBLP:conf/aaai/MalhotraS22}. Finally,
\UFO{} is the input format supported by the most recent implementation of
\textsc{FastWFOMC}\footnote{\url{https://github.com/jan-toth/FastWFOMC.jl}}. The
notation we use to refer to each logic is standard in the case of \Ctwo{} and
\UFO{}~\cite{tóth2024complexityweightedfirstordermodel} and redefined to be more
specific in the case of \FO{}. All three logics are function-free, and domains
are always assumed to be finite. As usual, we presuppose the \emph{unique name
  assumption}, which states that two constants are equal if and only if they are
the same constant~\cite{DBLP:books/aw/RN2020}.

@@

@@

% (In practice, we typically only specify the size of each domain.)

@@

While this work focuses on FOMC, we still define the weighted variant of the
problem as Skolemization relies on weights even for unweighted FOMC.

\todo[inline]{Even though we are evaluating the algorithms on FOMC benchmarks,
  for sentences with existential quantifiers, computing FOMC using \Cranetwo{}
  requires the use of WFOMC. For such sentences, Skolemization introduces
  predicates with non-unary weights that must be accounted for to compute the
  correct model count.}

@@

@@

We write $\expr{}$ for an arbitrary algebraic expression. In the context of
algebra, a \emph{constant} is a non-negative integer. Likewise, a
\emph{variable} can either be a parameter of a function or a variable introduced
through summation, such as $i$ in the expression $\sum_{i=1}^{n} \expr$. A
\emph{function call} is $f(x_{1}, \dots, x_{n})$ (or $f(\mathbf{x})$ for short),
where $f$ is an $n$-ary function, and each $x_{i}$ is an algebraic expression
consisting of variables and constants. A (function) \emph{signature} is function
call that contains only variables. Given two function calls $f(\mathbf{x})$ and
$f(\mathbf{y})$, we say that $f(\mathbf{y})$ \emph{matches} $f(\mathbf{x})$ if
$x_{i} = y_{i}$ whenever $x_{i}, y_{i} \in \mathbb{N}_{0}$. An \emph{equation}
is $f(\mathbf{x}) = \expr{}$, where $f(\mathbf{x})$ is a signature.

>>>>

FOMC algorithms are diverse, with approaches ranging from \emph{first-order
  knowledge compilation} (FOKC) to cell
enumeration~\cite{DBLP:conf/uai/BremenK21}, local
search~\cite{DBLP:journals/pvldb/NiuRDS11}, and Monte Carlo
sampling~\cite{DBLP:journals/cacm/GogateD16}. Among these, FOKC-based algorithms
are particularly prominent, transforming formulas into structured
representations such as circuits or graphs. Even when multiple algorithms are
able to solve the same instance, FOKC algorithms are known to find
polynomial-time solutions, where the polynomial has a lower degree compared to
other approaches~\cite{DBLP:conf/kr/DilkasB23}. The recently developed ability
of a FOKC algorithm to formulate solutions in terms of recursive
functions~\cite{DBLP:conf/kr/DilkasB23} is also noteworthy as the only other
proposed alternative is to guess recursive
relations~\cite{DBLP:conf/ilp/BarvinekB0ZK21}. Notable examples of FOKC
algorithms include \textsc{ForcLift}~\cite{DBLP:conf/ijcai/BroeckTMDR11} and its
successor \textsc{Crane}~\cite{DBLP:conf/kr/DilkasB23}.

@@

\todo[inline]{Adjust the introduction to the new structure (with more
  subsections)}

@@

\subsection{The Three Logics of FOMC}\label{sec:threelogics}

% \begin{table*}[t]
%   \centering
%   \begin{tabular}{llclll}
%     \toprule
%     Logic & Sorts & Constants & Variables & Quantifiers & Additional atoms\\
%     \midrule
%     \FO & one or more & \cmark & unlimited & $\forall$, $\exists$ & $x = y$\\
%     \Ctwo & one & \xmark & two & $\forall$, $\exists$, $\exists^{= k}$, $\exists^{\le k}$, $\exists^{\ge k}$ & ---\\
%     \UFO & one & \xmark & two & $\forall$ & $|P| = m$\\
%     \bottomrule
%   \end{tabular}
%   \caption{A comparison of the three logics used in FOMC\@. The
%     2\textsuperscript{nd}--5\textsuperscript{th} columns refer to: the number of
%     sorts, support for constants, the maximum number of variables, and supported
%     quantifiers, respectively. The last column lists supported atoms in addition
%     to those of the form $P(\mathbf{t})$ for a predicate $P/n$ and an $n$-tuple
%     of terms $\mathbf{t}$. Here: $k$ and $m$ are non-negative integers, with the
%     latter depending on the domain size, $P$ represents a predicate, and $x$ and
%     $y$ are terms. }\label{tbl:logics}
% \end{table*}

@@

There are three first-order logics commonly used in FOMC: \FO{}, \Ctwo{}, and
\UFO{}. First, \FO{} is the input format for
\textsc{ForcLift}\footnote{\url{https://github.com/UCLA-StarAI/Forclift}} and
its extensions
\textsc{Crane}\footnote{\url{https://doi.org/10.5281/zenodo.8004077}} and
\Cranetwo{}. Second, \Ctwo{} is often used in the literature on
\textsc{FastWFOMC}\footnote{\url{https://github.com/jan-toth/FastWFOMC.jl}} and
related methods~\cite{DBLP:journals/jair/Kuzelka21,DBLP:conf/aaai/MalhotraS22}.
Finally, \UFO{} is the input format supported by the most recent implementation
of \textsc{FastWFOMC}~\cite{tóth2024complexityweightedfirstordermodel}. All
three logics are function-free, and domains are always assumed to be finite. As
usual, we presuppose the \emph{unique name assumption}, which states that two
constants are equal if and only if they are the same
constant~\cite{DBLP:books/aw/RN2020}.

@@

\subsection{First-Order Model Counting}

\todo[inline,caption={}]{
  \begin{itemize}
    \item Write a proper introductory paragraph
    \item Maybe a bit more (easy-to-understand) detail about FOMC?
  \end{itemize}
}

@@

@@

\begin{remark*}
  Even though this work focuses on FOMC, for sentences with existential
  quantifiers, computing FOMC using \Cranetwo{} requires the use of WFOMC. For
  such sentences, Skolemization (described in \cref{sec:crane}) introduces
  predicates with non-unary weights that must be accounted for to compute the
  correct model count.
\end{remark*}

@@

\subsection{First-Order Knowledge Compilation}\label{sec:crane}

\todo[inline]{It would be nice to give the necessary introduction to FOKC, or
  \textsc{Crane} (an intuitive explanation early in the paper)}

\begin{example}
  Skolemization transforms
  \begin{equation}\label[formula]{eq:skolemizationinitial}
    \forall x \in \Gamma\text{. }\exists y \in \Delta\text{. } P(x, y)
  \end{equation}
  into
  \begin{equation}\label[formula]{eq:skolemization}
    \begin{aligned}
      (\forall x \in \Gamma\text{. } &Z(x))\land{}\\
      (\forall x \in \Gamma\text{. } \forall y \in \Delta\text{. } &Z(x) \lor \neg P(x, y))\land{}\\
      (\forall x \in \Gamma\text{. } &S(x) \lor Z(x))\land{}\\
      (\forall x \in \Gamma\text{. } \forall y \in \Delta\text{. } &S(x) \lor \neg P(x, y)).
    \end{aligned}
  \end{equation}
  With suitable weights on the new predicates $S/1$ and $Z/1$, the WFOMC of
  \cref{eq:skolemization} is equal to the FOMC of
  \cref{eq:skolemizationinitial}.
\end{example}

@@

In this paper, we use both logical and algebraic constructs. While the rest of
\cref{sec:preliminaries} focused on the former, this section describes the
latter. We write $\expr{}$ for an arbitrary algebraic expression. In the context
of algebra, a \emph{constant} is a non-negative integer. Likewise, a
\emph{variable} can either be a parameter of a function or a variable introduced
through summation, such as $i$ in the expression $\sum_{i=1}^{n} \expr$. A
\emph{function call} is $f(x_{1}, \dots, x_{n})$ (or $f(\mathbf{x})$ for short),
where $f$ is an $n$-ary function, and each $x_{i}$ is an algebraic expression
consisting of variables and constants. A (function) \emph{signature} is function
call that contains only variables. Given two function calls $f(\mathbf{x})$ and
$f(\mathbf{y})$, we say that $f(\mathbf{y})$ \emph{matches} $f(\mathbf{x})$ if
$x_{i} = y_{i}$ whenever $x_{i}, y_{i} \in \mathbb{N}_{0}$. An \emph{equation}
is $f(\mathbf{x}) = \expr{}$, where $f(\mathbf{x})$ is a signature.
